---
title: 'EDS 232 Lab 3: Reserve Planning'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lab 3: Reserve Planning

## R Packages

```{r}
if (!require("librarian")){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  assertthat, BiocManager, dplyr, gridExtra, here, mapview, 
  prioritizr, prioritizrdata, 
  raster, remotes, rgeos, rgdal, scales, sf, sp, stringr,
  units)
if (!require("lpsymphony")){
  BiocManager::install("lpsymphony")
  library(lpsymphony)
}
```

## Data Setup

```{r}
dir_data <- here("data/prioritizr")
pu_shp   <- file.path(dir_data, "pu.shp")
pu_url   <- "https://github.com/prioritizr/massey-workshop/raw/main/data.zip"
pu_zip   <- file.path(dir_data, basename(pu_url))
vegetation_tif <- file.path(dir_data, "vegetation.tif")

dir.create(dir_data, showWarnings = F, recursive = T)
if (!file.exists(pu_shp)){
  download.file(pu_url, pu_zip)
  unzip(pu_zip, exdir = dir_data)
  dir_unzip   <- file.path(dir_data, "data")
  files_unzip <- list.files(dir_unzip, full.names = T)
  file.rename(
    files_unzip, 
    files_unzip %>% str_replace("prioritizr/data", "prioritizr"))
  unlink(c(pu_zip, dir_unzip), recursive = T)
}
```

## Data

### Data Import

```{r}
# import planning unit data
pu_data <- as(read_sf(pu_shp), "Spatial")

# format columns in planning unit data
pu_data$locked_in <- as.logical(pu_data$locked_in)
pu_data$locked_out <- as.logical(pu_data$locked_out)

# import vegetation data
veg_data <- stack(vegetation_tif)
```

### Planning Unit Data

```{r}
# print a short summary of the data
print(pu_data)
```

```{r}
# plot the planning unit data
plot(pu_data)
```

```{r}
# plot an interactive map of the planning unit data
mapview(pu_data)
```

```{r}
# print the structure of object
str(pu_data, max.level = 2)
```

```{r}
# print the class of the object
class(pu_data)
```

```{r}
# print the slots of the object
slotNames(pu_data)
```

```{r}
# print the coordinate reference system
print(pu_data@proj4string)
```

```{r}
# print number of planning units (geometries) in the data
nrow(pu_data)
```

```{r}
# print the first six rows in the data
head(pu_data@data)
```

```{r}
# print the first six values in the cost column of the attribute data
head(pu_data$cost)
```

```{r}
# print the highest cost value
max(pu_data$cost)
```

```{r}
# print the smallest cost value
min(pu_data$cost)
```

```{r}
# print average cost value
mean(pu_data$cost)
```

```{r}
# plot a map of the planning unit cost data
spplot(pu_data, "cost")
```

```{r}
# plot an interactive map of the planning unit cost data
mapview(pu_data, zcol = "cost")
```

**How many planning units are in the planning unit data?**

*There are 516 planning units in the planning unit data set.*

**What is the highest cost value?**

*The highest cost value is 47.24 million Australian dollars.*

**Is there a spatial pattern in the planning unit cost values (hint: use plot to make a map)?**

*Yes. Lower costs tend to be on the east side, medium costs on the west side, and high costs on the north side.*

### Vegetation Data

```{r}
# print a short summary of the data
print(veg_data)
```

```{r}
# plot a map of the 20th vegetation class
plot(veg_data[[20]])
```

```{r}
# plot an interactive map of the 20th vegetation class
mapview(veg_data[[20]])
```

```{r}
# print number of rows in the data
nrow(veg_data)
```

```{r}
# print number of columns  in the data
ncol(veg_data)
```

```{r}
# print number of cells in the data
ncell(veg_data)
```

```{r}
# print number of layers in the data
nlayers(veg_data)
```

```{r}
# print  resolution on the x-axis
xres(veg_data)
```

```{r}
# print resolution on the y-axis
yres(veg_data)
```

```{r}
# print spatial extent of the grid, i.e. coordinates for corners
extent(veg_data)
```

```{r}
# print the coordinate reference system
print(veg_data@crs)
```

```{r}
# print a summary of the first layer in the stack
print(veg_data[[1]])
```

```{r}
# print the value in the 800th cell in the first layer of the stack
print(veg_data[[1]][800])
```

```{r}
# print the value of the cell located in the 30th row and the 60th column of
# the first layer
print(veg_data[[1]][30, 60])
```

```{r}
# calculate the sum of all the cell values in the first layer
cellStats(veg_data[[1]], "sum")
```

```{r}
# calculate the maximum value of all the cell values in the first layer
cellStats(veg_data[[1]], "max")
```

```{r}
# calculate the minimum value of all the cell values in the first layer
cellStats(veg_data[[1]], "min")
```

```{r}
# calculate the mean value of all the cell values in the first layer
cellStats(veg_data[[1]], "mean")
```


**What part of the study area is the 13th vegetation class found in (hint: make a map)? For instance, is it in the south-eastern part of the study area?**

```{r}
plot(veg_data[[13]])
```


*The 13th vegetation class can be found in the northeast and eastern part of the study area.*

**What proportion of cells contain the 12th vegetation class?**

```{r}
veg12_prop <- cellStats(veg_data[[12]], "mean")
```

*The proportion of cells that contain the 12th vegetation class is `r veg12_prop'.*

**Which vegetation class is the most abundant (i.e. present in the greatest number of cells)?**

**

## Gap Analysis

### Feature Abundance

```{r}
# create prioritizr problem with only the data
p0 <- problem(pu_data, veg_data, cost_column = "cost")

# print empty problem,
# we can see that only the cost and feature data are defined
print(p0)

# calculate amount of each feature in each planning unit
abundance_data <- feature_abundances(p0)

# print abundance data
print(abundance_data)
```

```{r}
# note that only the first ten rows are printed,
# this is because the abundance_data object is a tibble (i.e. tbl_df) object
# and not a standard data.frame object
print(class(abundance_data))
```

```{r}
# we can print all of the rows in abundance_data like this
print(abundance_data, n = Inf)
```

```{r}
# add new column with feature abundances in km^2
abundance_data$absolute_abundance_km2 <-
  (abundance_data$absolute_abundance * prod(res(veg_data))) %>%
  set_units(m^2) %>%
  set_units(km^2)

# print abundance data
print(abundance_data)
```

```{r}
# calculate the average abundance of the features
mean(abundance_data$absolute_abundance_km2)
```

```{r}
# plot histogram of the features' abundances
hist(abundance_data$absolute_abundance_km2, main = "Feature abundances")
```

```{r}
# find the abundance of the feature with the largest abundance
max(abundance_data$absolute_abundance_km2)
```

```{r}
# find the name of the feature with the largest abundance
abundance_data$feature[which.max(abundance_data$absolute_abundance_km2)]
```

**What is the median abundance of the features (hint: median)?**

```{r}
median_abundance <- median(abundance_data$absolute_abundance_km2)
```

*The median abundance of features is `r median_abundance`.*

**What is the name of the feature with smallest abundance?**

```{r}
smallest_abundance_feature <- abundance_data$feature[which.min(abundance_data$absolute_abundance_km2)]
```

*The feature with the smallest abundance is `r smallest_abundance_feature`.*

**How many features have a total abundance greater than 100 km^2 (hint: use sum(abundance_data$absolute_abundance_km2 > set_units(threshold, km^2) with the correct threshold value)?**

```{r}
over_100km2 <- sum(abundance_data$absolute_abundance_km2 > set_units(100, km^2))
```

*There are `r over_100km2` features that have a total abundance greater than 100km^2.*

### Feature Representation

```{r}
# create column in planning unit data with binary values (zeros and ones)
# indicating if a planning unit is covered by protected areas or not
pu_data$pa_status <- as.numeric(pu_data$locked_in)

# calculate feature representation by protected areas
repr_data <- eval_feature_representation_summary(p0, pu_data[, "pa_status"])

# print feature representation data
print(repr_data)
```

```{r}
# add new column with the areas represented in km^2
repr_data$absolute_held_km2 <-
  (repr_data$absolute_held * prod(res(veg_data))) %>%
  set_units(m^2) %>%
  set_units(km^2)

# print representation data
print(repr_data)
```

**What is the average proportion of the features held in protected areas (hint: use mean(table$relative_held) with the correct table name)?**

```{r}
avg_prop_protected <- mean(repr_data$relative_held)
```

*The average proportion of features held in protected areas is `r avg_prop_protected`.*

**If we set a target of 10% coverage by protected areas, how many features fail to meet this target (hint: use sum(table$relative_held >= target_value) with the correct table name)?**

```{r}
miss_target_num <- sum(repr_data$relative_held >= 0.10)
miss_target_num
```

*There are `r miss_target_num` features that fail to meet the 10% coverage by protected areas target.*

**If we set a target of 20% coverage by protected areas, how many features fail to meet this target?**

```{r}
miss_target_num20 <- sum(repr_data$relative_held >= 0.20)
miss_target_num20
```

*There are `r miss_target_num20` features that fail to meet the 20% coverage by protected areas target.*

**Is there a relationship between the total abundance of a feature and how well it is represented by protected areas (hint: plot(abundance_data$absolute_abundance ~ repr_data$relative_held))?**

```{r}
plot(abundance_data$absolute_abundance ~ repr_data$relative_held)
```

*I don't see any major correlation between total abundance of a feature and how well it is represented by protected areas.*

## Spatial Prioritizations

### Starting Out Simple

```{r}
# print planning unit data
print(pu_data)
```

```{r}
# make prioritization problem
p1_rds <- file.path(dir_data, "p1.rds")
if (!file.exists(p1_rds)){
  p1 <- problem(pu_data, veg_data, cost_column = "cost") %>%
        add_min_set_objective() %>%
        add_relative_targets(0.05) %>% # 5% representation targets
        add_binary_decisions() %>%
        add_lpsymphony_solver()
  saveRDS(p1, p1_rds)
}
p1 <- readRDS(p1_rds)

# print problem
print(p1)

# solve problem
s1 <- solve(p1)

# print solution, the solution_1 column contains the solution values
# indicating if a planning unit is (1) selected or (0) not
print(s1)
```

```{r}
# calculate number of planning units selected in the prioritization
eval_n_summary(p1, s1[, "solution_1"])
```

```{r}
# calculate total cost of the prioritization
eval_cost_summary(p1, s1[, "solution_1"])
```

```{r}
# plot solution
# selected = green, not selected = grey
spplot(s1, "solution_1", col.regions = c("grey80", "darkgreen"), main = "s1",
       colorkey = FALSE)
```

**How many planing units were selected in the prioritization? What proportion of planning units were selected in the prioritization?**

```{r}
eval_n_summary(p1, s1[, "solution_1"])
```

*15 planing units were selected in the prioritization.*

**Is there a pattern in the spatial distribution of the priority areas?**

*They appear to be pretty spread out.*

**Can you verify that all of the targets were met in the prioritization (hint: eval_feature_representation_summary(p1, s1[, "solution_1"]))?**

```{r}
feature_summary <- eval_feature_representation_summary(p1, s1[, "solution_1"])
sum(feature_summary$relative_held < 0.05)
```

*All of the `relative_held` are above 0.05, meaning that all of the targets were met in the prioritization.*

### Adding Complexity

```{r}
# plot locked_in data
# TRUE = blue, FALSE = grey
spplot(pu_data, "locked_in", col.regions = c("grey80", "darkblue"),
       main = "locked_in", colorkey = FALSE)
```

```{r}
# make prioritization problem
p2_rds <- file.path(dir_data, "p2.rds")
if (!file.exists(p2_rds)){
  p2 <- problem(pu_data, veg_data, cost_column = "cost") %>%
      add_min_set_objective() %>%
      add_relative_targets(0.05) %>%
      add_locked_in_constraints("locked_in") %>%
      add_binary_decisions() %>%
      add_lpsymphony_solver()
  saveRDS(p2, p2_rds)
}
p2 <- readRDS(p2_rds)

# print problem
print(p2)

# solve problem
s2 <- solve(p2)

# plot solution
# selected = green, not selected = grey
spplot(s2, "solution_1", col.regions = c("grey80", "darkgreen"), main = "s2",
       colorkey = FALSE)
```

```{r}
# make prioritization problem
p3_rds <- file.path(dir_data, "p3.rds")
if (!file.exists(p3_rds)){
  p3 <- problem(pu_data, veg_data, cost_column = "cost") %>%
    add_min_set_objective() %>%
    add_relative_targets(0.1) %>%
    add_locked_in_constraints("locked_in") %>%
    add_binary_decisions() %>%
    add_lpsymphony_solver()
  saveRDS(p3, p3_rds)
}
p3 <- readRDS(p3_rds)

# print problem
print(p3)

# solve problem
s3 <- solve(p3)

# plot solution
# selected = green, not selected = grey
spplot(s3, "solution_1", col.regions = c("grey80", "darkgreen"), main = "s3",
       colorkey = FALSE)
```

```{r}
# plot locked_out data
# TRUE = red, FALSE = grey
spplot(pu_data, "locked_out", col.regions = c("grey80", "darkred"),
       main = "locked_out", colorkey = FALSE)
```

```{r}
# make prioritization problem
p4_rds <- file.path(dir_data, "p4.rds")
if (!file.exists(p4_rds)){
  p4 <- problem(pu_data, veg_data, cost_column = "cost") %>%
    add_min_set_objective() %>%
    add_relative_targets(0.1) %>%
    add_locked_in_constraints("locked_in") %>%
    add_locked_out_constraints("locked_out") %>%
    add_binary_decisions() %>%
    add_lpsymphony_solver()
  saveRDS(p4, p4_rds)
}
p4 <- readRDS(p4_rds)
```

```{r}
# print problem
print(p4)

# solve problem
s4 <- solve(p4)

# plot solution
# selected = green, not selected = grey
spplot(s4, "solution_1", col.regions = c("grey80", "darkgreen"), main = "s4",
       colorkey = FALSE)
```

**What is the cost of the planning units selected in s2, s3, and s4?**

```{r}
eval_cost_summary(p2, s2[, "solution_1"])
eval_cost_summary(p3, s3[, "solution_1"])
eval_cost_summary(p4, s4[, "solution_1"])
```


*The cost for planning units in s2 is $6600.09 mil, s3 is $6669.91 mil, and s4 is $6711.58 mil.*

**How many planning units are in s2, s3, and s4?**

```{r}
eval_n_summary(p2, s2[, "solution_1"])
eval_n_summary(p3, s3[, "solution_1"])
eval_n_summary(p4, s4[, "solution_1"])
```

*The number planning units in s2 is 205, s3 is 211, and s4 is 212.*

**Do the solutions with more planning units have a greater cost? Why (or why not)?**

*Yes they do, because generally the higher number of planning units would mean higher cost.*

**Why does the first solution (s1) cost less than the second solution with protected areas locked into the solution (s2)?**

**

**Why does the third solution (s3) cost less than the fourth solution solution with highly degraded areas locked out (s4)?**

**

### Penalizing Fragmentation

```{r}
# make prioritization problem
p5_rds <- file.path(dir_data, "p5.rds")
if (!file.exists(p5_rds)){
  p5 <- problem(pu_data, veg_data, cost_column = "cost") %>%
    add_min_set_objective() %>%
    add_boundary_penalties(penalty = 0.001) %>%
    add_relative_targets(0.1) %>%
    add_locked_in_constraints("locked_in") %>%
    add_locked_out_constraints("locked_out") %>%
    add_binary_decisions() %>%
    add_lpsymphony_solver()
  saveRDS(p5, p5_rds)
}
p5 <- readRDS(p5_rds)

# print problem
print(p5)

# solve problem,
# note this will take a bit longer than the previous runs
s5 <- solve(p5)

# print solution
print(s5)
```

```{r}
# plot solution
# selected = green, not selected = grey
spplot(s5, "solution_1", col.regions = c("grey80", "darkgreen"), main = "s5",
       colorkey = FALSE)
```

**What is the cost the fourth (s4) and fifth (s5) solutions? Why does the fifth solution (s5) cost more than the fourth (s4) solution?**

**

**Try setting the penalty value to 0.000000001 (i.e. 1e-9) instead of 0.001. What is the cost of the solution now? Is it different from the fourth solution (s4) (hint: try plotting the solutions to visualize them)? Is this is a useful penalty value? Why (or why not)?**

**

**Try setting the penalty value to 0.5. What is the cost of the solution now? Is it different from the fourth solution (s4) (hint: try plotting the solutions to visualize them)? Is this a useful penalty value? Why (or why not)?**

**
